NDNS multilevel LCA CW2CB2 or CW2CB3

### 2018-08-26 paper/Chapters/Chapter2.tex
grammarly checked

Detailed dairy checking was performed to code and convert the food consumption into energy and nutrients intake. Intakes of nutrients were calculated from the food consumption records using a specially adapted Nutrient Databank \parencite{smithers1993maff}, which was originally developed by the Ministry of Agriculture, Fisheries and Food (MAFF) for the Dietary and Nutritional Survey of British Adults. Further details of data
coding and editing are outlined in Appendix A of the NDNS official reports \parencite{NDNSofficial}. Specifically, the main variables that we adopted in the current analysis were defined as:

@@

Time across a typical survey day was divided into 7 time slots in the dietary diary of NDNS RP: 6 am to 9 am, 9 am to 12 noon, 12 noon to 2 pm, 2 pm to 5 pm, 5 pm to 8 pm, 8 pm to 10 pm, and 10 pm to 6 am next morning. To produce a sequence of discrete responses regarding the carbohydrate intake we are interested, the energy consumption within each time slot over the four days of survey for each participant were calculated. The percentages of energy that contributed by carbohydrate within each time slot were then estimated. Since we planned to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, the responses were then dichotomised according to the carbohydrate contribution to the energy intake at cut-off value of 50\%, i.e. if within a time slot where there is any energy intake occured, carbohydrate consumption was categorised into whether it's energy contribution was lower or higher/equal to 50\% of total energy intake within that time slot. Consequently, for each day of the recording, there were 7 data points generated by the diary, each data point included one of the following responses:

@@

\begin{itemize}
	\item Randomly select PSUs from the Postcode Address File; 
	\item Randomly select 27 or 28 addresses in that postcode area; 
	\item Randomly select one household at that address; 
	\item Selected addresses were randomly allocated to one of two groups to determine whether an adult (aged 19 years or older) and a child (aged 1.5 to 18 years), or a child only, were selected for interview.
\end{itemize}
\vspace{-0.6cm}
%-----------------------------------
%	SUBSECTION 2
%-----------------------------------
\subsection{Response rates}\vspace{-0.3cm}

@@

It is necessary to apply weighting factors to the data collected in the NDNS RP for two reasons: to remove any bias in the observed results which may be due to differences in the probability of households and individuals being selected to take part; and to attempt to reduce differential non-response bias by age, sex and geographical region.

@@

\begin{enumerate}
	\item An overall selection weight, which is the product of the address, dwelling unit, catering (household) unit, and individual selection weights, was generated to correct for the unequal selection probabilities. These weights are the inverse of the selection probabilities at each level of the random sampling process and they can be used to compensate for differences in the chance of selection of an individual.
	\item An iterative procedure was used to adjust the selection weights until the distribution of the weighted sample matched that of the population for age-group, sex and geographical region. Population distributions were taken from the mid-year population estimates \parencite{OfficeofNationalStat}. 
\end{enumerate}

Another two sets of weights were generated to correct for differential non-response (either due to refusal or inability) to 1) nurse visit, and 2) giving a blood sample. Response rates to the nurse visit among those completed a dietary diary was approximately 75\%, to blood sample in adults were 51\%, 57\%, and 50\% for years 1 to 4, 5 to 6, and 7 to 8, respectively. In creating the nurse/blood sample weight, a logistic regression model was used by the NDNS RP study team to model the relationship between response to nurse visit/giving blood sample and a set of predictor variables (socio-demographic, participant and catering/household unit characteristics). The model generated a predicted probability for each participant, which is the probability would agree to a nurse visit/provide blood sample, given the characteristics of the individual and the household unit. Participants with characteristics associated with non-response were under-represented in the sample and therefore receive a low predicted probability. The inverse of these predicted probabilities were used as a set of non-response weights so that participants with a low predicted probability got a larger weight, increasing their representation in the sample. Then the nurse/blood sample weights were re-scaled so that the sum of the weights equalled the number of participants who had a nurse visit/who provided a blood sample. The final nurse/blood weights should therefore make the sample participants representative of all eligible persons in the population.

Further details of the weighting system developed by the NDNS RP are described in the Appendix B of the reports published by Public Health England (PHE) \parencite{bates2014national,roberts2018national,NDNSofficial}.\vspace{-0.6cm}

@@

Computer assisted personal interviews were conducted for the selected individuals by trained interviewers to collect background information on smoking habits (current, ex-smokers, and never), ethnicity (white, non-white), education level (lower than degree/degree or above level), living with a partner or not and other socio-demographic variables. Participants also had their height, weight, blood pressure, waist circumferences (WC) measured by the nurses.

Specifically, blood pressure was measured in a sitting position using an automated, validated machine, the Omron HEM907, after a five-minute rest. The means of second and third readings, taken at one minute intervals, were used in the current report. Hypertension was defined as with systolic blood pressure of 140 mmHg or above, and/or diastolic blood pressure of 90 mmHg or above, and/or taking any medication specifically to reduce blood pressure.

@@

Blood samples were stored at 4 $^\circ$C, and sent directly by post to the Department of Haematology and Department of Clinical Biochemistry and Immunology, Addenbrooke's Hospital, Cambridge within two hours of their collection. Serum samples were obtained by centrifugation of the coagulated blood sample. Serum total, High Density Lipoprotein (HDL) and Low Density Lipoprotein (LDL) cholesterol, triglycerides (TG), fasting blood glucose, haemoglobin A1C were measured. A1C value of 6.5\% was used as the cut off point for diagnosing diabetes.

@@

Ethical approval for the survey was obtained form the Oxfordshire A Research Ethics Committee. The letters of approval for the original submission and subsequent substantial amendments, together with approved documents, were sent to all Local Research Ethics Committees covering areas where fieldwork was being conducted. Research governance approval was sought for all participating NHS laboratories and obtained where required by the Research and Development Committee for each laboratory. Ethical approval for the current project was obtained from the MSc Research Ethics Committee of London School of Hygiene \& Tropical Medicine (LSHTM MSc Ethics Ref: 15624).

@@

Latent class analysis is a statistical technique that identifies categorical latent (unobserved) class variables on the basis of observed categorical variables \parencite{collins2010latent}. It belongs to the family of latent variable models, and is directly analogous to the factor analysis model. The major difference is that the latent variable in LCA is categorical, not continuous as in factor analysis. The basic assumptions in LCA are independent observations, and local independence, as shown in the fundamental expression of a typical LCA model: \vspace{-0.8cm}

\begin{equation}
P(U_{i1} = s_1, U_{i2} = s_2, \cdots, U_{ik} = s_k) = \sum_{t=t}^{T}P(C_i = t)\prod_{k = 1}^{K}P(U_{ik} = s_k | C_i = t)
\label{LCA}
\end{equation}\vspace{-0.7cm}

@@

Multilevel latent class analysis accounts for the nested structure of the data by allowing latent class intercepts to vary across level 2 units and thereby examining if and how level 2 units influence the level 1 latent classes. These random intercepts allow the probability of membership in a particular level 1 (observation days) latent class to vary across level 2 units (e.g., here in the current context are the individuals). Essentially this allows the probability that an observation day will belong to a
particular day level latent class to vary across individual levels.\vspace{-0.5cm}

@@

Proposed by Vermunt \parencite{Vermunt, vermunt2008latent} and Asparouhov and Muth\'en \parencite{muthen2009multilevel},  a traditional, parametric approach can be applied using a logistic regression model. For example, let's assume that there are two types of observation days in the dietary survey--high and low carbohydrate eating days. In an unconditional logistic regression model, the probability of the outcome (i.e. an observed high carbohydrate eating day vs. a low carbohydrate eating day) is constant within individual level, which means for each person throughout his/her survey there is some probability of following a high carbohydrate eating day. A random effect considers the individuals (level 2) to be drawn from the adult population in the UK, and the probability of the outcome (i.e. high carbohydrate eating days) across individuals is considered to be a random variable \parencite{snijders2011multilevel}.

@@

\begin{equation}
\begin{aligned}
\text{logit}[P(C_{ij} = 1)] & = \beta_{0j} + \beta_{1}x_{ij} \;\;\;\;\;\;\;\;\; \textbf{(day level)}  \\
\beta_{0j} & = \gamma_0 + \gamma_1 w_j + u_{0j} \; \textbf{(individual level)} \\ 
\Rightarrow P(C_{ij} = 1) & = \frac{\exp{(\gamma_0 + \beta_{1}x_{ij} + \gamma_1 w_j + u_{0j})}}{1 + \exp{(\gamma_0 + \beta_{1}x_{ij} + \gamma_1 w_j + u_{0j})}} \\
\end{aligned}
\label{randomLCA}
\end{equation}
\vspace{-0.3cm}

@@

\begin{itemize}
	\item $\text{logit}{(x)} = \log{\frac{x}{1-x}}$, where $x \in [0,1]$;
	\item $P(C_{ij} = 1)$ as the probability that the randomly selected $i$th observation day of $j$th individual is a high carbohydrate eating day;
	\item $\beta_{0j}$ as the random intercept, for outcome $C_{ij} = 1$; 
	\item the random deviation of the individuals $u_{0j}$ are assumed be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $x_{ij}, w_j$ is the predictors for day level (weekdays or weekends) and individual level, such as age, and sex.
\end{itemize}

Same framework can be used to consider random effects in an LCA model, but instead of saying that $C_{ij}$ is either low or high carbohydrate eating days as if we already know, it is now replaced by a latent variable $G_{ij}$ which indicates the typologies of carbohydrate eating patterns. Then we can use the day level data to assess the log-odds of belonging to $k$th type of carbohydrate eating pattern on a specific day of survey, and we allow the log-odds to vary across individuals. Therefore, for some persons the log-odds of having a $k$th type of carbohydrate eating pattern during the survey can be high, but for the other persons, the log-odds of following the $k$th type of carbohydrate eating pattern can be low.

If the day level LCA model (carbohydrate eating temporal pattern typologies) is best defined by $T (T \geqslant 2)$ latent classes, then $T-1$ random intercept will be specified by a two-level multinomial logistic regression model. Similar to the typical LCA models, the latent class variable in a MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each time slots, throughout three/four consecutive days of their survey period). Considering the latent class indicators are indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-1cm}

@@

\begin{itemize}
	\item $ U_{ijk} $ represents the response of carbohydrate eating (one of the following: not eating any food, $< 50\%$ of the energy, or $\geqslant 50\%$ of the energy) on $i$th day of the survey ($i \in (1,2,3,4)$) in $j$th individual at the $k$th time slot of the day ($k \in (1, 2, 3, \cdots, 7)$);
	\item $G_{ij}$ denotes the latent class membership for $j$th individuals on $i$th day of the survey, the total number of day level latent class is $T$;
%	\item A specific latent class is referred to as $t$, and the total number of level 1 latent classes is denoted by $T$;
	\item $P(U_{ijk} = s_k|C_{ij} = t)$ is the probability of a specific response pattern, conditional on membership in latent class $t$.
\end{itemize}

@@

Since the parametric approach discussed above can be extremely computationally demanding \parencite{van2008using, vermunt2008latent}, an alternative approach is using a non-parametric MLCA \parencite{davidian2008growth}. In this approach, separate latent class models are specified for level 1 (observation days) and level 2 (individuals). Similar with the parametric MLCA approach, there are $T-1$ random intercepts, where $T$ is the number of level 1 latent classes. However, rather than assuming the random intercepts following a normal distribution, the non-parametric MLCA assumes a multinomial (discrete) distribution of the level 2 latent classes. This approach is less computationally demanding compared with the parametric approach. These level 2 (individual) latent classes reflect differences in the probability of belonging to a specific day level latent class, so that individuals that have observation days with similar probabilities for the level 1 latent classes will be grouped together. The non-parametric MLCA model can be defined as follows:

@@

\begin{itemize}
	\item $CB_j$ is individual level latent class membership for $j$th individual;
	\item $\gamma_{tm}$ is day level and individual level indicators. 
\end{itemize}

@@

\begin{itemize}
	\item Firstly, we ignored the multilevel structure of the data and estimated a series of traditional LC models to determine the number of classes at the observational day-level;
	\item Next, a series of MLCA models were fitted to account for the multilevel structrure of the data. In these models, the number of day-level classes was based on the best fitting LCA model from the first step, and the LCA model at the individual level was estimated to identify the number of individual level latent classes;
	\item Thirdly, when number of individual level latent classes is defined based on the previous stage, number of day-level classes was modified (one class lower and one class higher than in the second step) to see the effect of changing level 1 classes and confirm the best fitting model.
\end{itemize}

The number of classes in level 1 was determined by 1) the evaluation of model fit indices, including the Bayesian information criterion (BIC), and entropy which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compares $q$ vs. $q-1$ classes models, where $q$ is the number of latent classes and most importantly, 3) pattern interpretability. In the steps of performing multilevel LCA, where LMR-LRT is not available, same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in observation day level and individual level. MLCA models were fitted in Mplus 7.4 \parencite{muthen2005mplus}, the Mplus syntax and outputs are shown in \textbf{Appendix \ref{AppendixB}}. \vspace{-0.3cm}

@@

\subsection{Characteristics of day level latent classes and individual level latent classes}\vspace{-0.3cm}

Day level latent classes identified by the first step of MLCA were tabulated according to day of week and also whether the diary was recorded during weekends or not. A contingency table giving the frequency of responses across the 7 time slots of the survey days was produced. Descriptive statistics for the dietary day level recordings according to the latent class memberships were presented. Pearson $\chi^2$ test was used to compare the distribution of categorical variables. One-way Analysis of Variance (ANOVA) was used to compare the means across the multiple groups for continuous variables.

Person level point estimates and 95\% confidence intervals (CIs) were determined by applying individual, nurse visiting, and blood sample weights accordingly which account for the probability of participant selection and the clustered survey design. Descriptive statistics for sample characteristics were presented as weighted means (95\% CI) or weighted percentages (95\%CI). After examining the distribution of the data, the following variables were log-transformed to improve normality: fasting blood glucose, A1C, TC, LDL, HDL, TG, and average physical activity duration per day. Weighted geometric means (95\% CI) were used for all log-transformed variables.

To see whether there is any temporal pattern for food intake eating could also be defined at individual level, weighted estimates of nutrients consumption across the 7 time slots of the day were calculated for each individual level latent class. Contributions (\%) of the average energy intake within time slots were evaluated by determining the percentages of energy coming from carbohydrate, fat, protein, and alcohol intake.

@@

\subsection{Association between individual level latent classes and the prevalence of hypertension, and measurements of obesity}\vspace{-0.3cm}

Associations between individual level carbohydrate eating classes and hypertension (yes/no), body mass index (BMI, kg/m\textsuperscript{2}), and waist circumference (WC, cm) were explored in men and women separately. Point estimates of weighted means and proportions and 95\%CI of the characteristics were determined by applying either nurse visiting weights (for outcomes of hypertension, BMI, and WC) or blood sample weights (for diagnosis of DM) accordingly. Similarly, \textit{F} tests (for continuous variables) and adjusted Pearson $\chi^2$ tests (for categorical variables) were used to determine sex-specific differences by hypertension status, and BMI categories.

@@

Survey-designed logistic regression models (for hypertension), and linear regression models (for WC, BMI), were used to test for associations between latent classes of carbohydrate eating patterns and hypertension, BMI, and WC, in the NDNS RP sample, separately. Since diabetic participants might or might not modify their carbohydrate eating habits, we also fitted all the above mentioned regression models restricted to those without diabetes.

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statisitical analyses conducted above, i.e. those are associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Those are strongly related with the outcomes but may not associated with carbohydrate eating groups may reduce the standard errors and so improve the precisions are also considered in the linear regression models. 
	\item Confounding and/or interaction effect from each of the potential factors were checked one by one. Interaction effect were tested using the adjusted Wald test testing whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in condition of the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using the adapted \code{svylogitgof} command in Stata \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardized Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), assumption of independent observations is violated as soon as we weighted the sample. General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are not available as well. Outliers, leverage, and Cook's distance cannot be checked either, however, participants with extreme weightings (if exist) were checked by removing them and refit the models as a sensitivity analysis.
	\item Since under survey design data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was also used to  decide whether quadratic and cubic terms of continuous variables were necessary in improving the fitting of model \parencite{pregibon1980goodness}. 
\end{enumerate}

>>>>

Detailed dairy checking was performed to code and convert the food consumption into energy and nutrients intake. Intakes of nutrients were calculated from the food consumption records using a specially adapted Nutrient Databank \parencite{smithers1993maff}, which was originally developed by the Ministry of Agriculture, Fisheries, and Food (MAFF) for the Dietary and Nutritional Survey of British Adults. Further details of data
coding and editing are outlined in Appendix A of the NDNS official reports \parencite{NDNSofficial}. Specifically, the main variables that we adopted in the current analysis were defined as:

@@

Time across a typical survey day was divided into 7 time slots in the dietary diary of NDNS RP: 6 am to 9 am, 9 am to 12 noon, 12 noon to 2 pm, 2 pm to 5 pm, 5 pm to 8 pm, 8 pm to 10 pm, and 10 pm to 6 am next morning. To produce a sequence of discrete responses regarding the carbohydrate intake we are interested, the energy consumption within each time slot over the four days of survey for each participant were calculated. The percentages of total energy intake that contributed by carbohydrate within each time slot were then estimated. Since we planned to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, the responses were then dichotomised according to the carbohydrate contribution to the energy intake at cut-off value of 50\%, i.e. if within a time slot where there is any energy intake occurred, carbohydrate consumption was categorised into whether it's energy contribution was lower or higher/equal to 50\% of total energy intake within that time slot. Consequently, for each day of the recording, there were 7 data points generated by the diary, each data point included one of the following responses:

@@

\begin{itemize}
	\item Randomly select PSUs from the Postcode Address File; 
	\item Randomly select 27 or 28 addresses in that postcode area; 
	\item Randomly select one household at that address; 
	\item Selected addresses were randomly allocated to one of two groups to determine whether an adult (aged 19 years or older) and a child (aged 1.5 to 18 years) or a child only, were selected for interviews.
\end{itemize}
\vspace{-0.6cm}
%-----------------------------------
%	SUBSECTION 2
%-----------------------------------
\subsection{Response rates}\vspace{-0.3cm}

@@

It is necessary to apply weighting factors to the data collected in the NDNS RP for two reasons: to remove any bias in the observed results which may be due to differences in the probability of households and individuals being selected to take part; and to attempt to reduce differential non-response bias by age, sex, and geographical region.

@@

\begin{enumerate}
	\item An overall selection weight, which is the product of the address, dwelling unit, catering (household) unit, and individual selection weights, was generated to correct for the unequal selection probabilities. These weights are the inverses of the selection probabilities at each level of the random sampling process and they can be used to compensate for differences in the chance of selection of an individual.
	\item An iterative procedure was used to adjust the selection weights until the distribution of the weighted sample matched that of the population for age-group, sex, and geographical region. Population distributions were taken from the mid-year population estimates \parencite{OfficeofNationalStat}. 
\end{enumerate}

Another two sets of weights were generated to correct for differential non-response (either due to refusal or inability) to 1) nurse visit and 2) giving a blood sample. Response rates to the nurse visit among those completed a dietary diary was approximately 75\%, to the blood sample in adults were 51\%, 57\%, and 50\% for years 1 to 4, 5 to 6, and 7 to 8, respectively. In creating the nurse/blood sample weight, a logistic regression model was used by the NDNS RP study team to model the relationship between response to nurse visit/giving a blood sample and a set of predictor variables (socio-demographic, participant and catering/household unit characteristics). The model generated a predicted probability for each participant, which is the probability would agree to a nurse visit/provide a blood sample, given the characteristics of the individual and the household unit. Participants with characteristics associated with non-response were under-represented in the sample and therefore receive a low predicted probability. The inverses of these predicted probabilities were used as a set of non-response weights so that participants with a low predicted probability got a larger weight, increasing their representation in the sample. Then the nurse/blood sample weights were re-scaled so that the sum of the weights equalled the number of participants who had a nurse visit/who provided a blood sample. The final nurse/blood weights should, therefore, make the sample participants representative of all eligible persons in the population.

Further details of the weighting system developed by the NDNS RP are described in Appendix B of the reports published by Public Health England (PHE) \parencite{bates2014national,roberts2018national,NDNSofficial}.\vspace{-0.6cm}

@@

Computer-assisted personal interviews were conducted for the selected individuals by trained interviewers to collect background information on smoking habits (current, ex-smokers, and never), ethnicity (white, non-white), education level (lower than degree/degree or above level), living with a partner or not and other socio-demographic variables. Participants also had their height, weight, blood pressure, waist circumferences (WC) measured by the nurses.

Specifically, blood pressure was measured in a sitting position using an automated, validated machine, the Omron HEM907, after a five-minute rest. The means of second and third readings, taken at one-minute intervals, were used in the current report. Hypertension was defined as with systolic blood pressure of 140 mmHg or above, and/or diastolic blood pressure of 90 mmHg or above, and/or taking any medication specifically to reduce blood pressure.

@@

Blood samples were stored at 4 $^\circ$C, and sent directly by post to the Department of Haematology and Department of Clinical Biochemistry and Immunology, Addenbrooke's Hospital, Cambridge within two hours of their collection. Serum samples were obtained by centrifugation of the coagulated blood sample. Serum total, High-Density Lipoprotein (HDL) and Low-Density Lipoprotein (LDL) cholesterol, triglycerides (TG), fasting blood glucose, haemoglobin A1C were measured. A1C value of 6.5\% was used as the cut off point for diagnosing diabetes.

@@

Ethical approval for the survey was obtained from the Oxfordshire A Research Ethics Committee. The letters of approval for the original submission and subsequent substantial amendments, together with approved documents, were sent to all Local Research Ethics Committees covering areas where fieldwork was being conducted. Research governance approval was sought for all participating NHS laboratories and obtained where required by the Research and Development Committee for each laboratory. Ethical approval for the current project was obtained from the MSc Research Ethics Committee of the London School of Hygiene \& Tropical Medicine (LSHTM MSc Ethics Ref: 15624).

@@

Latent class analysis is a statistical technique that identifies categorical latent (unobserved) class variables on the basis of observed categorical variables \parencite{collins2010latent}. It belongs to the family of latent variable models and is directly analogous to the factor analysis model. The major difference is that the latent variable in LCA is categorical, not continuous as in factor analysis. The basic assumptions in LCA are independent observations, and local independence, as shown in the fundamental expression of a typical LCA model: \vspace{-0.8cm}

\begin{equation}
P(U_{i1} = s_1, U_{i2} = s_2, \cdots, U_{ik} = s_k) = \sum_{t=t}^{T}P(C_i = t)\prod_{k = 1}^{K}P(\mathbf{U_{ik}} = \mathbf{s_k} | C_i = t)
\label{LCA}
\end{equation}\vspace{-0.7cm}

@@

Multilevel latent class analysis accounts for the nested structure of the data by allowing latent class intercepts to vary across level 2 units and thereby examining if and how level 2 units influence the level 1 latent classes. These random intercepts allow the probability of membership in a particular level 1 (observation days) latent class to vary across level 2 units (e.g., here in the current context are the individuals). Essentially this allows the probability that an observation day will belong to a
particular day-level latent class to vary across individual-levels.\vspace{-0.5cm}

@@

Proposed by Vermunt \parencite{Vermunt, vermunt2008latent} and Asparouhov and Muth\'en \parencite{muthen2009multilevel},  a traditional, parametric approach can be applied using a logistic regression model. For example, let's assume that there are two types of observation days in the dietary survey--high and low carbohydrate eating days. In an unconditional logistic regression model, the probability of the outcome (i.e. an observed high carbohydrate eating day vs. a low carbohydrate eating day) is constant within the individual-level, which means for each person throughout his/her survey there is some probability of following a high carbohydrate eating day. A random effect considers the individuals (level 2) to be drawn from the adult population in the UK, and the probability of the outcome (i.e. high carbohydrate eating days) across individuals is considered to be a random variable \parencite{snijders2011multilevel}.

@@

\begin{equation}
\begin{aligned}
\text{logit}[P(C_{ij} = 1)] & = \beta_{0j} + \beta_{1}x_{ij} \;\;\;\;\;\;\;\;\; \textbf{(day-level)}  \\
\beta_{0j} & = \gamma_0 + \gamma_1 w_j + u_{0j} \; \textbf{(individual-level)} \\ 
\Rightarrow P(C_{ij} = 1) & = \frac{\exp{(\gamma_0 + \beta_{1}x_{ij} + \gamma_1 w_j + u_{0j})}}{1 + \exp{(\gamma_0 + \beta_{1}x_{ij} + \gamma_1 w_j + u_{0j})}} \\
\end{aligned}
\label{randomLCA}
\end{equation}
\vspace{-0.3cm}

@@

\begin{itemize}
	\item $\text{logit}{(x)} = \log{\frac{x}{1-x}}$, where $x \in [0,1]$;
	\item $P(C_{ij} = 1)$ as the probability that the randomly selected $i$th observation day of the $j$th individual is a high carbohydrate eating day;
	\item $\beta_{0j}$ as the random intercept, for the outcome $C_{ij} = 1$; 
	\item the random deviation of the individuals $u_{0j}$ are assumed be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $x_{ij}, w_j$ is the predictors for day-level (weekdays or weekends) and individual-level, such as age, and sex.
\end{itemize}

The same framework can be used to consider random effects in an LCA model, but instead of saying that $C_{ij}$ is either low or high carbohydrate eating days as if we already know, it is now replaced by a latent variable $G_{ij}$ which indicates the typologies of carbohydrate eating patterns. Then we can use the day-level data to assess the log-odds of belonging to the $k$th type of carbohydrate eating pattern on a specific day of the survey, and we allow the log-odds to vary across individuals. Therefore, for some persons, the log-odds of having a $k$th type of carbohydrate eating pattern during the survey can be high, but for the other persons, the log-odds of following the $k$th type of carbohydrate eating pattern can be low.

If the day-level LCA model (carbohydrate eating temporal pattern typologies) is best defined by $T (T \geqslant 2)$ latent classes, then $T-1$ random intercept will be specified by a two-level multinomial logistic regression model. Similar to the typical LCA models, the latent class variable in a MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each time slots, throughout three/four consecutive days of their survey period). Considering the latent class indicators are indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-1cm}

@@

\begin{itemize}
	\item $ U_{ijk} $ represents the response of carbohydrate eating (one of the following: not eating any food, $< 50\%$ of the energy, or $\geqslant 50\%$ of the energy) on the $i$th day of the survey ($i \in (1,2,3,4)$) in $j$th individual at the $k$th time slot of the day ($k \in (1, 2, 3, \cdots, 7)$);
	\item $G_{ij}$ denotes the latent class membership for $j$th individuals on $i$th day of the survey, the total number of day-level latent class is $T$;
%	\item A specific latent class is referred to as $t$, and the total number of level 1 latent classes is denoted by $T$;
	\item $P(U_{ijk} = s_k|C_{ij} = t)$ is the probability of a specific response pattern, conditional on membership in latent class $t$.
\end{itemize}

@@

Since the parametric approach discussed above can be extremely computationally demanding \parencite{van2008using, vermunt2008latent}, an alternative approach is using a non-parametric MLCA \parencite{davidian2008growth}. In this approach, separate latent class models are specified for level 1 (observation days) and level 2 (individuals). Similar with the parametric MLCA approach, there are $T-1$ random intercepts, where $T$ is the number of level 1 latent classes. However, rather than assuming the random intercepts following a normal distribution, the non-parametric MLCA assumes a multinomial (discrete) distribution of the level 2 latent classes. This approach is less computationally demanding compared with the parametric approach. These level 2 (individual) latent classes reflect differences in the probability of belonging to a specific day-level latent class, so that individuals that have observation days with similar probabilities for the level 1 latent classes will be grouped together. The non-parametric MLCA model can be defined as follows:

@@

\begin{itemize}
	\item $CB_j$ is the individual-level latent class membership for the $j$th individual;
	\item $\gamma_{tm}$ is day-level and individual-level indicators. 
\end{itemize}

@@

\begin{itemize}
	\item Firstly, we ignored the multilevel structure of the data and estimated a series of traditional LC models to determine the number of classes at the observational day-level;
	\item Next, a series of MLCA models were fitted to account for the multilevel structure of the data. In these models, the number of day-level classes was based on the best fitting LCA model from the first step, and the LCA model at the individual-level was estimated to identify the number of individual-level latent classes;
	\item Thirdly, when the number of individual-level latent classes is defined based on the previous stage, the number of day-level classes was modified (one class lower and one class higher than in the second step) to see the effect of changing level 1 classes and confirm the best fitting model.
\end{itemize}

The number of classes in level 1 was determined by 1) the evaluation of model fit indices, including the Bayesian information criterion (BIC), and entropy which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compares $q$ vs. $q-1$ classes models, where $q$ is the number of latent classes and most importantly, 3) pattern interpretability. In the steps of performing multilevel LCA, where LMR-LRT is not available, same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in observation day-level and individual-level. MLCA models were fitted in Mplus 7.4 \parencite{muthen2005mplus}, the Mplus syntax and outputs are shown in \textbf{Appendix \ref{AppendixB}}. \vspace{-0.3cm}

@@

\subsection{Characteristics of day-level latent classes and individual-level latent classes}\vspace{-0.3cm}

day-level latent classes identified by the first step of MLCA were tabulated according to the day of week and also whether the diary was recorded during weekends or not. A contingency table giving the frequency of responses across the 7-time slots of the survey days was produced. Descriptive statistics for the dietary day-level recordings according to the latent class memberships were presented. Pearson $\chi^2$ test was used to compare the distribution of categorical variables. One-way Analysis of Variance (ANOVA) was used to compare the means across the multiple groups for continuous variables.

Person-level point estimates and 95\% confidence intervals (CIs) were determined by applying individual, nurse visiting, and blood sample weights accordingly which account for the probability of participant selection and the clustered survey design. Descriptive statistics for sample characteristics were presented as weighted means (95\% CI) or weighted percentages (95\%CI). After examining the distribution of the data, the following variables were log-transformed to improve normality: fasting blood glucose, A1C, TC, LDL, HDL, TG, and average physical activity duration per day. Weighted geometric means (95\% CI) were used for all log-transformed variables.

To see whether there is any temporal pattern for food intake eating could also be defined at the individual-level, weighted estimates of nutrients consumption across the 7-time slots of the day were calculated for each individual-level latent class. Contributions (\%) of the average energy intake within time slots were evaluated by determining the percentages of energy coming from carbohydrate, fat, protein, and alcohol intake.

@@

\subsection{Association between individual-level latent classes and the prevalence of hypertension, and measurements of obesity}\vspace{-0.3cm}

Associations between individual-level carbohydrate eating classes and hypertension (yes/no), body mass index (BMI, kg/m\textsuperscript{2}), and waist circumference (WC, cm) were explored in men and women separately. Point estimates of weighted means and proportions and 95\%CI of the characteristics were determined by applying either nurse visiting weights (for outcomes of hypertension, BMI, and WC) or blood sample weights (for diagnosis of DM) accordingly. Similarly, \textit{F} tests (for continuous variables) and adjusted Pearson $\chi^2$ tests (for categorical variables) were used to determine sex-specific differences by hypertension status, and BMI categories.

@@

Survey-designed logistic regression models (for hypertension), and linear regression models (for WC, BMI), were used to test for associations between latent classes of carbohydrate eating patterns and hypertension, BMI, and WC, in the NDNS RP sample, separately. Since diabetic participants might or might not modify their carbohydrate eating habits, we also fitted all the above-mentioned regression models restricted to those without diabetes.

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statistical analyses conducted above, i.e. those are associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Those are strongly related with the outcomes but may not associate with carbohydrate eating groups may reduce the standard errors and so improve the precisions are also considered in the linear regression models. 
	\item Confounding and/or interaction effect from each of the potential factors were checked one by one. Interaction effects were tested using the adjusted Wald test testing whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in the condition of the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using the adapted \code{svylogitgof} command in Stata \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardized Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), the assumption of independent observations is violated as soon as we weighted the sample. General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are not available either. Outliers, leverage, and Cook's distance cannot be checked either, however, participants with extreme weightings (if exist) were checked by removing them and refit the models as a sensitivity analysis.
	\item Since under survey designed data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was also used to  decide whether quadratic and cubic terms of continuous variables were necessary in improving fitting of the model \parencite{pregibon1980goodness}. 
\end{enumerate}
